A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-07-01 papers/contact/paper.tex
fix bug in simplification of white bear and fix bug in build
fix bugs in contact density (boy that was a pain)

The hard-sphere contact density---which we will define as the density
of hard spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).  For
a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$---is easily computed from the
Carnahan-Starling equation of state
\begin{align}
  g_{HS}(\sigma) &= \frac{1}{1-\eta} + \cdots
\end{align}
The contact density may be easily computed from a knowledge of the
contact theorem, which states that the pressure on any hard surface is
given by
\begin{align}
  p &= k_BT n_\textit{contact}
\end{align}
Since we are interested in the contact density at the hard-sphere
surface, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the free energy on hard sphere
radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
Since $\eta \equiv \frac{4\pi}{3} R^3$, we can work out that
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$.  The area which is relevant
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
  &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  &= n k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
As it turns out, this is the standard answer for the contact value of
the autocorrelation function! $\ddot\smile$

@@

  Here is the White Bear free energy functional for the hard-sphere
  fluid:~\cite{roth2002whitebear}
  \begin{align}
    A_{HS} &= \int \left\{
    -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
\end{align}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}
    %\left(
    %  \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    %\right)
    \\
    & \times \left(\frac{1}{n_3(1-n_3)^3} -\frac1{n_3^2(1-n_3)^2}  -
      \frac{1}{n_3^2(1-n_3)} - 2\frac{\ln(1-n_3)}{n_3^3}\right)
\end{align}
\end{widetext}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_0(\mathbf{r}')} &= -\ln(1-n_3)
    \\
    \beta\frac{\delta A_{HS}}{\delta n_1(\mathbf{r}')} &= \frac{n_2}{1-n_3}
    \\
    \beta\frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
      \frac{n_1}{1-n_3}
      + (n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        12\pi n_3^2(1-n_3)^2
      }
\end{align}
\begin{align}
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V1}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V2}}{1-n_3}
    \\
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V1}}{1-n_3}
      - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        36\pi n_3^2(1-n_3)^2
      }
  \end{align}

@@

Here is the White Bear free energy functional for the hard-sphere
fluid:~\cite{roth2002whitebear}
\begin{align}
  \frac{A_{HS}}{k_BT} &= \int \left\{
  -n_0 \ln\left( 1 - n_3\right)
  + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
    n_3 + (1-n_3)^2 \ln(1-n_3)
  }{
    36\pi n_3^2(1-n_3)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n^3 (4\pi R^2)^3 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n^3 4\pi (4\pi R^3)^2 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n\eta^2 4\pi 9 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{\eta}{1-\eta}
  + \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{\eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  + \ln(1-\eta)
  \right\} \\
  &= N \left\{
  \frac{\eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N
  \frac{\eta(1-\eta) + \eta
  }{
    (1-\eta)^2
  } \\
  &= N
  \frac{2\eta -\eta^2}{(1-\eta)^2}
\end{align}
This is odd, because Wikipedia says that for the Carnahan-Starling
equation of state the excess free energy is
\begin{align}
  A = Nk_BT\frac{4\eta -3\eta^2}{(1-\eta)^2}
\end{align}
\end{widetext}

>>>>

The hard-sphere contact density---which we will define as the density
of hard spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).  For
a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$---is easily computed from the
Carnahan-Starling equation of state
\begin{align}
  g_{HS}(\sigma) &= \frac{1}{1-\eta} + \cdots
\end{align}
The contact density may be easily computed from a knowledge of the
contact theorem, which states that the pressure on any hard surface is
given by
\begin{align}
  p &= k_BT n_\textit{contact}
\end{align}
Since we are interested in the contact density at the hard-sphere
surface, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the free energy on hard sphere
radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
Since $\eta \equiv \frac{4\pi}{3} R^3$, we can work out that
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$.  The area which is relevant
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
  &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  &= n k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
As it turns out, this is the standard answer for the contact value of
the autocorrelation function! $\ddot\smile$

@@

  Here is the White Bear free energy functional for the hard-sphere
  fluid:~\cite{roth2002whitebear}
  \begin{align}
    \beta A_{HS} &= \int \left\{
    -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
\end{align}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}
    %\left(
    %  \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    %\right)
    \\
    & \times \left(\frac{2}{n_3(1-n_3)^3} -\frac1{n_3^2(1-n_3)^2}  -
      \frac{1}{n_3^2(1-n_3)} - 2\frac{\ln(1-n_3)}{n_3^3}\right) \\
    \beta\frac{\partial A_{HS}}{\partial n_3}
    &\sim \frac{n}{1-\eta} + \frac{3n\eta}{(1-\eta)^2} +
      \frac{2n\eta}{(1-\eta)^3} -
      \frac{n}{(1-\eta)^2} -
      \frac{n}{1-\eta} -
      2n \frac{ \ln(1-\eta) }{ \eta } \\
    &= \frac{3n\eta}{(1-\eta)^2} +
      \frac{2n\eta}{(1-\eta)^3} -
      \frac{n}{(1-\eta)^2} -
      2n \frac{ \ln(1-\eta) }{ \eta } \\
    &= n\left(
      \frac{3\eta - 3 \eta^2}{(1-\eta)^3} +
      \frac{2\eta}{(1-\eta)^3} -
      \frac{1 - \eta}{(1-\eta)^3} -
      2 \frac{ \ln(1-\eta) }{ \eta }
      \right) \\
    &= n\left(
      \frac{6\eta - 3 \eta^2 - 1}{(1-\eta)^3} -
      2 \frac{ \ln(1-\eta) }{ \eta }
      \right)
\end{align}
\end{widetext}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_0(\mathbf{r}')} &= -\ln(1-n_3)
    \\
    \beta\frac{\delta A_{HS}}{\delta n_1(\mathbf{r}')} &= \frac{n_2}{1-n_3}
    \\
    \beta\frac{\partial A_{HS}}{\partial n_1}
    &\sim n \frac{4\pi R^2}{1-\eta}
\end{align}
\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
      \frac{n_1}{1-n_3}
      + (n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        12\pi n_3^2(1-n_3)^2
      }
\end{align}
\begin{align}
    \beta\frac{\partial A_{HS}}{\partial n_2}
    &\sim n\frac{R}{1-\eta} +
    \frac{4\pi}{3} R^4 n^2 \frac{\eta + (1-\eta)^2\ln(1-\eta)}{\eta^2(1-\eta)^2}
    \\
    &= n\frac{R}{1-\eta} +
    \frac{R n}{(1-\eta)^2}
    + Rn \frac{\ln(1-\eta)}{\eta}
    \\
    &= nR\left( \frac{1 + \eta^2  - 2\eta}{(1-\eta)^3} +
      \frac{1 - \eta}{(1-\eta)^3}
    + \frac{\ln(1-\eta)}{\eta}
    \right)
    \\
    &= nR\left( \frac{2 + \eta^2  - 3\eta}{(1-\eta)^3}
    + \frac{\ln(1-\eta)}{\eta}
    \right)
\end{align}
\begin{align}
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V1}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V2}}{1-n_3}
    \\
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V1}}{1-n_3}
      - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        36\pi n_3^2(1-n_3)^2
      }
  \end{align}

\begin{widetext}
\begin{align}
  \beta \frac{\partial A_{HS}}{\partial R} &\sim
    \beta \frac{\partial A_{HS}}{\partial n_3} n_2 +
    \beta \frac{\partial A_{HS}}{\partial n_2} 2 \frac{n_2}{R} +
    \beta \frac{\partial A_{HS}}{\partial n_1} n
  \\
  &=
    n_2n \left(
      \frac{6\eta - 3\eta^2 - 1}{(1-\eta)^3} -
      2\frac{\ln(1-\eta)}{\eta}
    \right) +
    n_2n \left( \frac{4+2\eta^2-6\eta}{(1-\eta)^3} + 2\frac{\ln(1-\eta)}{\eta} \right) +
    n_2n\frac{1}{1-\eta}
  \\
  &=
    n_2n \left(
      \frac{6\eta - 3\eta^2 - 1}{(1-\eta)^3} +
    \frac{4+2\eta^2-6\eta}{(1-\eta)^3} +
    \frac{1 - 2\eta + \eta^2}{(1-\eta)^3}
  \right)
  \\
  &=
    n_2n \frac{4 - 2\eta}{(1-\eta)^3}
\end{align}
\end{widetext}
Recall from Equation~\ref{eq:dAhsdR}, the standard Carnahan result is
\begin{align}
  \frac{dA_{HS}}{dR}
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R}
  \\ &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} n_2
\end{align}
which means that we miraculously come out with a correct answer!

@@

Here is the White Bear free energy functional for the hard-sphere
fluid:~\cite{roth2002whitebear}
\begin{align}
  \frac{A_{HS}}{k_BT} &= \int \left\{
  -n_0 \ln\left( 1 - n_3\right)
  + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
    n_3 + (1-n_3)^2 \ln(1-n_3)
  }{
    36\pi n_3^2(1-n_3)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n^2 (4\pi R^2)  (R)}{1-\eta}
  + n^3 (4\pi R^2)^3 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{3 n \eta}{1-\eta}
  + n^3 (4\pi R^2)^3 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{3 n \eta}{1-\eta}
  + n^3 4\pi (4\pi R^3)^2 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{3 n \eta}{1-\eta}
  + n\eta^2 4\pi 9 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{3 \eta}{1-\eta}
  + \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{3 \eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  + \ln(1-\eta)
  \right\} \\
  &= N \left\{
  \frac{3\eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N
  \frac{3\eta(1-\eta) + \eta
  }{
    (1-\eta)^2
  } \\
  &= N
  \frac{4\eta -3\eta^2}{(1-\eta)^2}
\end{align}
This matches the usual expression for the Carnahan-Starling equation
of state the excess free energy, which is
\begin{align}
  A = Nk_BT\frac{4\eta -3\eta^2}{(1-\eta)^2}
\end{align}
\end{widetext}
